---
title: "EDA"
output:
  word_document: default
  html_notebook: default
---

```{r}
OG <- read.csv("~/Data/CollegeTM2018.csv")

TM <- OG

library(lme4)
library(tidyverse)
```

Part 1

```{r}
#map hit types to run values from
#https://rfrey22.medium.com/collegiate-linear-weights-f0237cf40451
#this is our response variable

TM <- TM%>%
  filter(!PlayResult == "Undefined")%>%
  mutate(runval = ifelse(
    PlayResult == "Single", 0.94, ifelse(
      PlayResult == "Double", 1.34, ifelse(
        PlayResult == "Triple", 1.67, ifelse(
          PlayResult == "HomeRun", 2.08, 0)))))%>%
  select(runval, ExitSpeed, Angle, Direction, Distance,
         PitcherId, BatterId, Stadium)

summary(TM)
```

a)
```{r}
#There are a ton of pitches that did not result in batted balls or had no recorded Exit Speeds, so they will be removed.

TM.clean <- TM%>%
  filter(!is.na(ExitSpeed))

#we want the ID variables to be factors, not numeric
TM.clean$PitcherId = as.factor(TM.clean$PitcherId)
TM.clean$BatterId = as.factor(TM.clean$BatterId)

summary(TM.clean)
```

b)

```{r}
#Data Exploration

hist(TM.clean$runval)
hist(TM.clean$ExitSpeed)
hist(TM.clean$Angle)
hist(TM.clean$Direction)
hist(TM.clean$Distance)
table(OG$HitType[OG$HitType != "Undefined"])/ 
  length(OG$HitType[OG$HitType != "Undefined"])
```


```{r}
length(unique(TM.clean$Stadium))
length(unique(TM.clean$BatterId))
length(unique(TM.clean$PitcherId))

```


c) Pairs of Variables
```{r}
plot(TM.clean$runval, TM.clean$ExitSpeed)
plot(TM.clean$ExitSpeed, TM.clean$Angle)
plot(TM.clean$Distance, TM.clean$Angle)
plot(TM.clean$Angle, TM.clean$Direction)
```




PART 3: MODEL BUILDING


Center Variables
```{r}
TM.clean = TM.clean%>%
  mutate(ES.c = ExitSpeed - mean(ExitSpeed))
```


Fixed effects only model (NULL)

```{r}
mod1 <- lm(runval ~ ExitSpeed + Angle + Direction + Distance, data = TM.clean)
summary(mod1)
regclass::VIF(mod1)
AIC(mod1)
```
The model with only fixed effects had all 4 variables significant with no multicollinearity concerns.

Add interactions
```{r}
mod1a <- lm(runval ~ ES.c + Angle + Direction + Distance + 
              ES.c:Angle + ES.c:Direction + ES.c:Distance +
              Angle:Direction + Angle:Distance + Distance:Direction, data = TM.clean)
summary(mod1a)
#regclass::VIF(mod1a)
AIC(mod1a)
```
Once we add interaction terms, the model improves a lot and four of the interaction terms are significant. We will ignore these for now while we figure out the random structure of our model and add these four terms back in at the end.


RANDOM STRUCTURE
```{r}
# sample = sample(nrow(TM.clean), size = 10000)
# TM.clean_sample = TM.clean[sample,]

# model1 <- lmer(runval ~ I(ExitSpeed - mean(TM$ExitSpeed)) + Angle +
#                  Direction + Distance +
#     (1 | PitcherId) +  (1 | BatterId) + (1 | Stadium), data = TM.clean)

model1_ints <- lmer(runval ~ ExitSpeed + Angle + Direction + Distance + (1|BatterId), data = TM.clean)
summary(model1_ints)

AIC(model1_ints)

```
Allowing for random intercepts alone for BatterId does not help us and in fact increased AIC a bit


```{r}
mod2_ints2 <- lmer(runval ~ ExitSpeed + Angle + Direction + Distance + (1|BatterId/Stadium), data = TM.clean)

summary(mod2_ints2)

AIC(mod2_ints2)
```

Three level model with BatterId at level 2 and Stadium at level 3 was an improvement over the last model but not over the entirely fixed model.


```{r}
mod2_ints3 <- lmer(runval ~ ExitSpeed + Angle + Direction + Distance + (1|Stadium), data = TM.clean)

summary(mod2_ints3)

AIC(mod2_ints3)
```

A 2 level model with StadiumId random intercepts is our best random effects model yet, but still is not better than the fixed model.

```{r}


mod3 <- lmer(runval ~ ES.c + Angle + Direction + Distance + (1 + ES.c | Stadium), data = TM.clean)

summary(mod3)

AIC(mod3)
```

A 2 level model with stadiumId random intercepts and centered Exit Speed random slopes is our best model yet and is better than the fixed model by AIC

```{r}
mod3b <- lmer(runval ~ ES.c + Angle + Direction + Distance + (1 + ES.c | BatterId/Stadium), data = TM.clean)

summary(mod3b)

AIC(mod3b)
BIC(mod3b)
```

Again, best model yet with 3 levels (hits within batter within stadium) and allowing exit speed to vary at batter and stadium level.

```{r}
mod3c <- lmer(runval ~ ES.c + Angle + Direction + Distance + (1 + ES.c + Distance | BatterId/Stadium), data = TM.clean)

summary(mod3c)

AIC(mod3c)
BIC(mod3c)
```

Adding Distance random slopes at the batter and stadium level did not help our model by BIC (barely helped by AIC). Considering Exit Speed shouldn't vary much at the stadium level, we will only allow Exit Speed slopes to vary at the batter level while leaving Distance to vary at the batter and stadium level. We will not calculate the cross level correlations  (hence the syntax).

```{r}
mod3d <- lmer(runval ~ ES.c + Angle + Direction + Distance + (1 + ES.c | BatterId) + (-1 + Distance|BatterId/Stadium) +
                (1|Stadium) , data = TM.clean)

summary(mod3d)

AIC(mod3d)
BIC(mod3d)
```

Best model yet for sure.

See if adding Angle random slopes at batter level helps

```{r}
mod3e <- lmer(runval ~ ES.c + Angle + Direction + Distance + (1 + ES.c + Angle | BatterId) + (-1 + Distance|BatterId/Stadium) + (1|Stadium) , data = TM.clean)

summary(mod3e)

AIC(mod3e)
BIC(mod3e)
```

This is worse. Angle should not be allowed to vary at the batter level.

Try adding Direction at level 2 and 3

```{r}
mod3f <- lmer(runval ~ ES.c + Angle + Direction + Distance + (1 + ES.c | BatterId) + (-1 + Distance + Direction|BatterId/Stadium) + (1|Stadium) , data = TM.clean)

summary(mod3f)

AIC(mod3f)
BIC(mod3f)
```

Adding direction at level 2 and 3 was not helpful. We suspect just having Direction random slopes vary at the batter level will be better 

```{r}
mod3f2 <- lmer(runval ~ ES.c + Angle + Direction + Distance + 
                 (1 + ES.c + Direction| BatterId) + 
                 (-1 + Distance |BatterId/Stadium) + 
                 (1|Stadium), data = TM.clean)

summary(mod3f2)

AIC(mod3f2)
BIC(mod3f2)
```

This is our best model so far (likely final).

Rearrange terms for more parsimony?
```{r}
mod3h <- lmer(runval ~ ES.c + Angle + Direction + Distance + 
                (1 + Distance +  ES.c | BatterId) + 
                (1  + Direction|BatterId/Stadium) , data = TM.clean)

summary(mod3h)

AIC(mod3h)
BIC(mod3h)
```
This rearrangement of terms did not help the model and hurt it pretty badly. We are satisfied with the previous model as being our best. Now we will add those interaction terms back in.


Add significicant interactions back in from earlier

```{r}

mod4 <- lmer(runval ~ ES.c + Angle + Direction + Distance + 
              ES.c:Angle  + ES.c:Distance + Angle:Distance + Distance:Direction +
                 (1 + ES.c + Direction| BatterId) + 
                 (-1 + Distance |BatterId/Stadium) + 
                 (1|Stadium), data = TM.clean)

summary(mod4)
AIC(mod4)
```

This is by far our best model

Remove insignifcant terms (just ES.c)
```{r}
mod4a <- lmer(runval ~ Angle + Direction + Distance + 
              ES.c:Angle  + ES.c:Distance + Angle:Distance + Distance:Direction +
                 (1 + ES.c + Direction| BatterId) + 
                 (-1 + Distance |BatterId/Stadium) + 
                 (1|Stadium), data = TM.clean)

summary(mod4a)
AIC(mod4a)
```

This hurt the model. Our final model will be model 4.




